Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2023-12-21 ATACCC.tex
Fix references

Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient MCMC implementation of the NUTS algorithm (see \cref{E-backcalc:sec:MCMC}).
The first of these is using a continuous viral load trajectory rather than the more common piecewise linear model.
As NUTS uses gradient information, it is more efficient when the model is differentiable.
This leads to the following model for the viral load at time $t$, $v_i(t)$, in individual $i$ is shown in \cref{ATACCC:fig:viral-load-model} and given by:
\begin{align}
v_i(\tau) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(\tau-\tau_{\max,i})} + a_ie^{b_i(\tau-\tau_{max,i})}}. \label{ATACCC:eq:viral-load}
\end{align}
Time, $\tau$, is continuous and defined uniquely for each individual such that $\tau = 0$ at the time that individual $i$ was first tested.
$v_{\max,i}$ is the maximum viral load, which occurs at time $\tau_{\max,i}$.
As $\lvert \tau - \tau_{\max} \rvert$ becomes large, the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
% Before the peak, this is because $b(\tau-\tau_{\max,i})$ is small and hence $a\exp(b(\tau-\tau_{\max,i}))$ is negligible.
% After the peak, this is because $b\exp(-a(\tau-\tau_{\max,i}))$ is negligible.
This happens quickly for realistic values of the parameters.
Therefore, the log viral load is approximately piecewise linear (see \cref{ATACCC:fig:viral-load-model} for a representative viral load trajectory).
$\vec{\phi}_i = (v_{\max,i}, a_i, b_i)^T$ are the random effects for individual $i$.
$\tau_{\max,i}$ is inferred independently for each individual because the time from first test to peak is not a biological concept shared between individuals.
\begin{figure}
  \centering \includegraphics{ATACCC/typical_trajectory}
  \caption[Example viral load model trajectory.]{Example viral load trajectory produced by \textcite{hakkiOnset}'s viral load model, using realistic parameter values. The increase in log viral load is approximately linear with a gradient of $a_i$, here 5. The decrease is also linear, with gradient $-b_i$, here -1.4. The peak is at time $\tau_{\max,i}$, here 0, and log viral load $\log v_{\max,i}$, here = 15. The smoothing (lack of linearity) near the peak is visible. \label{ATACCC:fig:viral-load-model}}
  \todo[inline]{Add a few lines to example plot showing the effect of different parameters, move parameters to key}
\end{figure}

@@

\todo[inline]{insert a better explanation of $F_D$'s posterior}
For later use, a discretised version is preferred (as described in \cref{E-inc-prev}).
To calculate the discrete version of this distribution, $F_D$, a Monte Carlo integration is used with the following relation.
\begin{align}
  F_D(d)
  &= \prob(D_i \leq d + 0.5) \\
  &= \E_{D_i}(\I(D_i \leq d + 0.5)) 
\end{align}
where $\I$ is the indicator function.
Here, $D_i$ is drawn conditional on the population-level parameters $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
Equivalent forms of $F_D$, such as the survival and hazard functions, can be calculated using the relations in \cref{E-backcalc:sec:definitions}.
I found 100,000 draws produced stable estimates of the tail of $F_D$.

\section{Results} \label{ATACCC:sec:results-discussion}

@@

The $F_D$ estimates for the two groups are in \cref{ATACCC:fig:duration}.
In \cref{ATACCC:sec:results-discussion}, the two models gave very similar results.
Therefore, I show results only from the modified model (which had better convergence).
\begin{figure}
  \centering \includegraphics{ATACCC/duration}
  \caption[Duration of PCR positivity.]{Estimated distribution (survival function) of duration of PCR positivity for the two populations. The vaccinated population have shorter durations of PCR positivity.  \label{ATACCC:fig:duration}}
\end{figure}

>>>>

Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient MCMC implementation of the NUTS algorithm (see \cref{E-inc-prev:sec:MCMC}).
The first of these is using a continuous viral load trajectory rather than the more common piecewise linear model.
As NUTS uses gradient information, it is more efficient when the model is differentiable.
This leads to the following model for the viral load at time $t$, $v_i(t)$, in individual $i$ is shown in \cref{ATACCC:fig:viral-load-model} and given by:
\begin{align}
v_i(\tau) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(\tau-\tau_{\max,i})} + a_ie^{b_i(\tau-\tau_{max,i})}}. \label{ATACCC:eq:viral-load}
\end{align}
Time, $\tau$, is continuous and defined uniquely for each individual such that $\tau = 0$ at the time that individual $i$ was first tested.
$v_{\max,i}$ is the maximum viral load, which occurs at time $\tau_{\max,i}$.
As $\lvert \tau - \tau_{\max} \rvert$ becomes large, the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
% Before the peak, this is because $b(\tau-\tau_{\max,i})$ is small and hence $a\exp(b(\tau-\tau_{\max,i}))$ is negligible.
% After the peak, this is because $b\exp(-a(\tau-\tau_{\max,i}))$ is negligible.
This happens quickly for realistic values of the parameters.
Therefore, the log viral load is approximately piecewise linear (see \cref{ATACCC:fig:viral-load-model} for a representative viral load trajectory).
$\vec{\phi}_i = (v_{\max,i}, a_i, b_i)^T$ are the random effects for individual $i$.
$\tau_{\max,i}$ is inferred independently for each individual because the time from first test to peak is not a biological concept shared between individuals.
\begin{figure}
  \centering \includegraphics{ATACCC/typical_trajectory}
  \caption[Example viral load model trajectory.]{Example viral load trajectory produced by \textcite{hakkiOnset}'s viral load model, using realistic parameter values. The increase in log viral load is approximately linear with a gradient of $a_i$, here 5. The decrease is also linear, with gradient $-b_i$, here -1.4. The peak is at time $\tau_{\max,i}$, here 0, and log viral load $\log v_{\max,i}$, here = 15. The smoothing (lack of linearity) near the peak is visible. \label{ATACCC:fig:viral-load-model}}
  \todo[inline]{Add a few lines to example plot showing the effect of different parameters, move parameters to key}
\end{figure}

@@

\todo[inline]{insert a better explanation of $F_D$'s posterior}
For later use, a discretised version is preferred (as described in \cref{E-inc-prev}).
To calculate the discrete version of this distribution, $F_D$, a Monte Carlo integration is used with the following relation.
\begin{align}
  F_D(d)
  &= \prob(D_i \leq d + 0.5) \\
  &= \E_{D_i}(\I(D_i \leq d + 0.5)) 
\end{align}
where $\I$ is the indicator function.
Here, $D_i$ is drawn conditional on the population-level parameters $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
Equivalent forms of $F_D$, such as the survival and hazard functions, can be calculated using the relations in \cref{E-inc-prev:sec:definitions}.
I found 100,000 draws produced stable estimates of the tail of $F_D$.

\section{Results} \label{ATACCC:sec:results}

@@

The $F_D$ estimates for the two groups are in \cref{ATACCC:fig:duration}.
In \cref{ATACCC:sec:results}, the two models gave very similar results.
Therefore, I show results only from the modified model (which had better convergence).
\begin{figure}
  \centering \includegraphics{ATACCC/duration}
  \caption[Duration of PCR positivity.]{Estimated distribution (survival function) of duration of PCR positivity for the two populations. The vaccinated population have shorter durations of PCR positivity.  \label{ATACCC:fig:duration}}
\end{figure}

### 2023-12-21 biology-data.tex
Fix references

\emph{Intermittent negatives} are the clearest example of a false negative.
An intermittent negative is when an individual tests negative but tested positive previously and subsequently.
Intermittent negatives are stripped out when creating the dataset used for all duration analyses in this chapter (as described in \cref{E-episode-def}), but demonstrate that false negatives do occur.

Whether a future positive is part of the same episode of a reinfection is not always trivial; I rely on a process developed previously by Sarah Walker, see \cref{E-episode-def}.
However, up until the emergence of the Omicron variant, reinfections are rare, especially in a short time frame, and hence the tricky cases are rare until this time (late 2021).

>>>>

\emph{Intermittent negatives} are the clearest example of a false negative.
An intermittent negative is when an individual tests negative but tested positive previously and subsequently.
Intermittent negatives are stripped out when creating the dataset used for all duration analyses in this chapter, but demonstrate that false negatives do occur.

% Whether a future positive is part of the same episode of a reinfection is not always trivial; I rely on a process developed previously by Sarah Walker, see \cref{E-episode-def}.
However, up until the emergence of the Omicron variant, reinfections are rare, especially in a short time frame, and hence the tricky cases are rare until this time (late 2021).

### 2023-12-21 cis-perfect-testing.tex
Fix references

Inference was implemented in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
Convergence was assessed using Rhat and ESS as described in \cref{E-ATACCC:sec:convergence}\todo{add ref to place explaining diagnostics}, and all runs checked for divergent transitions.

@@

I have assumed constant and independent incidence (a priori).
Prevalence in the CIS data was approximately constant over the period of interest (see \cref{E-intro:fig:prevalence}).
A sensitivity analysis, where an epidemic in exponential growth or decline was simulated (not shown), showed minimal impact on survival estimates.
An incorrect assumption of constant incidence can lead to biased estimates~\autocite{degruttolaAnalysis}.
Therefore, the assumption could be important in other contexts, especially if the incidence is changing rapidly.

>>>>

Inference was implemented in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
Convergence was assessed using Rhat and ESS as described in \cref{E-inc-prev:sec:MCMC}, and all runs checked for divergent transitions.

@@

I have assumed constant and independent incidence (a priori).
Prevalence in the CIS data was approximately constant over the period of interest (see\todo{ref prev constant over period}).
A sensitivity analysis, where an epidemic in exponential growth or decline was simulated (not shown), showed minimal impact on survival estimates.
An incorrect assumption of constant incidence can lead to biased estimates~\autocite{degruttolaAnalysis}.
Therefore, the assumption could be important in other contexts, especially if the incidence is changing rapidly.

### 2023-12-21 incidence-prevalence.tex
Fix references

The proposal distribution $q$ is most commonly a (possibly multivariate) Normal distribution centred on $\psi$; this scheme is used in \cref{E-SEIR}.
Tuning the proposal distribution to propose ``good'' moves is a key part of MCMC.
The ideal proposal distribution is the posterior of interest, however, this is impossible because that is what MCMC is trying to estimate~\autocite[296]{gelmanBDA}.
One alternative is \emph{adaptive} algorithms which tune the proposal distribution based on previous draws in an attempt to approximate the posterior; I use such a scheme in \cref{E-SEIR:sec:MCMC}.

>>>>

The proposal distribution $q$ is most commonly a (possibly multivariate) Normal distribution centred on $\psi$; this scheme is used in \cref{E-SEIR}.
Tuning the proposal distribution to propose ``good'' moves is a key part of MCMC.
The ideal proposal distribution is the posterior of interest, however, this is impossible because that is what MCMC is trying to estimate~\autocite[296]{gelmanBDA}.
One alternative is \emph{adaptive} algorithms which tune the proposal distribution based on previous draws in an attempt to approximate the posterior; I use such a scheme in \cref{E-SEIR:sec:inference}.

### 2023-12-21 introduction.tex
Fix references

I start this chapter by introducing a generic framework for mechanistic modelling of infectious diseases (\cref{E-SEIR:sec:transmission-generic}).
% This framework is then applied to SARS-CoV-2 in \cref{E-SEIR:sec:transmission-application}.
It describes the \emph{transmission model}, corresponding to the infection process.
The transmission model needs to be linked to observations; in this chapter, that will be the CIS prevalence data.
The \emph{observation model}, introduced in \cref{E-SEIR:sec:observation}, is this link; it incorporates both the prevalence process and the observation process.
These two components form the mathematical model, which I then fit to data, requiring specification of an inference process, parameterisation, and priors (\cref{E-SEIR:sec:inference-implementation}).
This final formulation uses the transmission model from \textcite{birrellRealtime}, but a novel observation model.
I then use a simulation study to check this recovers the parameters successfully (\cref{E-SEIR:sec:sim-study}).
Having shown that the model can recover the parameters, I then apply it to the CIS data (\cref{E-SEIR:sec:application}).
Finally, I discuss the results, limitations, and extensions of this work (\cref{E-SEIR:sec:discussion}).

Notation in this chapter differs from the rest of this thesis.
I adopt conventions and notation familiar to the infectious disease modelling field, although the field's notation is not consistent between sources.
In particular, upper or lower case no longer signifies whether a variable is random or not, but this is clear from the context.

>>>>

% I start this chapter by introducing a generic framework for mechanistic modelling of infectious diseases (\cref{E-SEIR:sec:transmission-generic}).
% % This framework is then applied to SARS-CoV-2 in \cref{E-SEIR:sec:transmission-application}.
% It describes the \emph{transmission model}, corresponding to the infection process.
% The transmission model needs to be linked to observations; in this chapter, that will be the CIS prevalence data.
% The \emph{observation model}, introduced in \cref{E-SEIR:sec:observation}, is this link; it incorporates both the prevalence process and the observation process.
% These two components form the mathematical model, which I then fit to data, requiring specification of an inference process, parameterisation, and priors (\cref{E-SEIR:sec:inference-implementation}).
% This final formulation uses the transmission model from \textcite{birrellRealtime}, but a novel observation model.
% I then use a simulation study to check this recovers the parameters successfully (\cref{E-SEIR:sec:sim-study}).
% Having shown that the model can recover the parameters, I then apply it to the CIS data (\cref{E-SEIR:sec:application}).
% Finally, I discuss the results, limitations, and extensions of this work (\cref{E-SEIR:sec:discussion}).
